# Reproducible Research: Peer Assessment 1, Joseph 

## Loading and preprocessing the data

```{r echo=TRUE}
date()
library(plyr)

basedir <- file.path("C:","Users","browse","reproducibleresearch","RepData_PeerAssessment1")
datadir <- file.path(basedir,"activity")
DA <- read.csv(file.path(datadir,"activity.csv"),as.is=TRUE) # data frame
dim(DA) # verify 17568 intervals in the data set
head(DA,15) # examine the data
DA <- DA[order(DA$date,DA$interval),] # make sure that we are ordered!

# looks like interval is concatonated hour minutes, so have to separate:
DA$hour <- floor(DA$interval/100)
DA$minute <- DA$interval - 100*DA$hour
DA$weekday <- weekdays(as.Date(DA$date))
DA$weekend <- ifelse(DA$weekday %in% c("Saturday","Sunday"),"Weekend","Weekday")

```

## What is mean total number of steps taken per day?
By linearity of sums, mean daily steps is the sum of the mean in each interval: 

```{r}
# get mean across all days for each interval, ignoring NAs: 

# function to get interval characteristics (median and mean) of steps by date:
getDMM <- function(DA, append = ""){
  DMean <- ddply(DA,"interval",function(D){mean(D$steps,na.rm=TRUE)})
  names(DMean) <- c("interval",paste("meansteps",append,sep=""))

  DMedian <- ddply(DA,"interval",function(D){median(D$steps,na.rm=TRUE)})
  names(DMedian) <- c("interval",paste("mediansteps",append,sep=""))

  DMM <- merge(DMean,DMedian)
  return(DMM)
  
}

DMM <- getDMM(DA)
DMM.Weekend <- getDMM( subset(DA, weekend=="Weekend"), ".Weekend")
DMM.Weekday <- getDMM( subset(DA, weekend=="Weekday"), ".Weekday")
DMM <- merge(DMM,DMM.Weekend)
DMM <- merge(DMM,DMM.Weekday)


# impute missing values, substituting NAs with median values across all days:
DA$stepsB <- DA$steps
DB <- ddply(DA,"date",
  function(D){
  D$stepsB[is.na(D$stepsB)] <- DMM$mediansteps[is.na(D$stepsB)] # rely on order of intervals 
  return(D)
  })



DD <- ddply(DB,"date",function(D){sum(D$stepsB)}) # summarize steps each day
names(DD) <- c("date","dailysteps")
DD$weekday <- weekdays(as.Date(DD$date))
DD$weekend <- ifelse(DD$weekday %in% c("Saturday","Sunday"),"Weekend","Weekday")

hist(DD$dailysteps,xlab = "Total steps", main = "Histogram of steps per day")
abline(v=median(DD$dailysteps),col='red',lty=2)
abline(v=mean(DD$dailysteps),col='blue',lty=2)
grid()
legend('topright',lty=2,col=c('red','blue'),legend=c("Median","Mean"))

summary(DD$dailysteps)
MeanStepsPerDay <- mean(DD$dailysteps)
MedianStepsPerDay <- median(DD$dailysteps)
MaxInterval <- DMM$interval[DMM$meansteps==max(DMM$meansteps)]


```
* Median steps Per Day: `r MedianStepsPerDay`
* Mean steps Per Day: `r MeanStepsPerDay`

## What is the average daily activity pattern?
* Interval with max mean steps is `r MaxInterval` 

Here's asummary plot of mean and median across all days:

```{r fig.height=5, fig.width=5}
plot(DMM$interval,pmax(DMM$mediansteps,DMM$meansteps),type='n',
  main = "Daily steps vs. interval, all days",xlab="interval",ylab="steps")
lines(DMM$interval,DMM$mediansteps,col='red')
lines(DMM$interval,DMM$meansteps,col='blue')
abline(h=MaxInterval)
grid()
legend('topright',lty=1,col=c('red','blue'),legend=c("Median","Mean"))



```

## Imputing missing values
```{r }
Nnas <- sum(is.na(DA$steps))
Nrows <- dim(DA)[1] 
```
* In the original data, there were `r Nnas` NAs out of `r Nrows` intervals.

* Imputation substituted median value across all days for each interval, in place of NAs.

## Are there differences in activity patterns between weekdays and weekends?
Yes, see the plot differences!
```{r fig.height=5, fig.width=10}

par(mfrow = c(1,2))

plot(DMM$interval,pmax(DMM$mediansteps,DMM$meansteps),type='n',
  main = "Daily steps vs. interval, Weekday",xlab="interval",ylab="steps")
lines(DMM$interval,DMM$mediansteps.Weekday,col='red')
lines(DMM$interval,DMM$meansteps.Weekday,col='blue')
grid()
legend('topright',lty=1,col=c('red','blue'),legend=c("Median","Mean"))

plot(DMM$interval,pmax(DMM$mediansteps,DMM$meansteps),type='n',
  main = "Daily steps vs. interval, Weekend",xlab="interval",ylab="steps")
lines(DMM$interval,DMM$mediansteps.Weekend,col='red')
lines(DMM$interval,DMM$meansteps.Weekend,col='blue')
grid()
legend('topright',lty=1,col=c('red','blue'),legend=c("Median","Mean"))
```
